Architectures for workflow integration in science gateways

### 2016-04-19 paper.tex
Wrote section 2.1

In the last decade, the e-Science workflow community has developed
high-level workflow systems to help developers access distributed
infrastructures such as grids and web services, resulting in tools
such as Askalon, Hyperflow, MOTEUR, Pegasus, Swift, Taverna, Triana,
VizTrails, WS-PGRADE/gUSE, etc. Such workflow engines usually describe
applications in a specific language that offers operators for parallel
computing, visual description and edition, links with domain-specific
tool repositories, etc. Descriptions and experiments conducted with
e-Science workflow engines were published in journals such as Future
Generation Computer Systems and conference venues such as
WORKS.

At the same time, specific toolboxes were emerging in various
scientific domains to facilitate interactions between different
software components. In neuroimaging for instance, tools such as
Nipype, PSOM, SPM and FSL provide abstractions and functions to define
processes that handle the data flow between other processes. Soon,
such tools were interfaced to computing systems, in particular
clusters: they were extended to create cluster tasks, handle their
dependencies, and execute them on clusters. For instance, FSL can
launch tasks on SGE through its \texttt{fsl\_sub} tool, Nipype
\todo{...}, PSOM \todo{...}, and SPM \todo{...}. Although distributed
infrastructures are hardly handled, such domain-specific tools have to
be called workflow engines. They represent a tremendous opportunity
for science gateways to leverage existing tools and
applications. Whether these should be combined with e-Science workflow
engines is an open question.

A workflow engine is defined as a software that submits jobs to a
cluster, grid or cloud. Workflows may be expressed in any language,
including scripts. Several workflow engines may be combined in the
same science gateway. Workflow: a process that submits tasks to the
infrastructure. FSL tools, say Feat, can be considered as workflows
when they are configured to submit tasks to clusters using fslsub. A
workflow consists of activities. We are talking only about concrete
workflows with a configuration (see terstyansky et al 2014 "enabling
scientific workflow sharing...")

  Workflow engine: executes workflows, i.e. process their dependencies
  and create computing tasks. Might transfer data too. A workflow
  engine is not supposed to be aware of the science gateway.

@@

@@

\todo{Maybe this is not relevant}

@@

\item Workflow integration: consists in adding a new workflow to the
  system so that users can execute it. It is triggered by an
  administrator of the science gateway and it results in an interface,
  for instance a web form, where users can enter the parameters of the
  workflow to be executed. The interaction has two aspects:
  (\texttt{a$_1$}) the programs used in the workflow are installed on
  the infrastructure, which may or may not require administrative
  privileges on the infrastructure; (\texttt{a$_2$}) the workflow is
  configured in the science gateway so that it becomes available to
  users. Note that integrating a workflow is not the same process as
  integrating a workflow \emph{engine}.
\item Task control: operations to manage tasks on the infrastructure,
  including: authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires to deal with the
  heterogeneous batch managers and meta-schedulers that might be
  available on the infrastructure. When the infrastructure is a grid
  or a cloud, it is for instance implemented using libraries such as
  SAGA, DRMAA, OCCI and similar initiatives \todo{refs needed}.
\item Data control: operations to manage data on the infrastructure,
  such as: upload, download, deletion, browsing, replication, caching,
  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer data across the infrastructure. The
  infrastructure might offer various data storage backends with
  heterogeneous interfaces. Tools and services such as
  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
  can be used to homogeneize these interfaces.
\item Workflow control: operations to execute a workflow with an
  engine, including: workflow submission, monitoring, termination,
  etc. Workflow control can be coarse-grained (a.k.a. black box) or
  fine-grained (white box). In a coarse-grained model, workflow
  activities are masked and the user only has a global view of the
  workflow execution. In a fine-grained model, user is exposed to the
  workflow topology, i.e. to the outputs of the individual tasks,
  their statuses and so on. 
\item Sub-task control: operations used by tasks to submit sub-tasks
  on the infrastructure, including: submission, monitoring,
  termination, deletion, etc. Sub-task control is similar to
  interaction \texttt{b}, except that information about the parent
  task which submitted a sub-task is usually available and used for
  additional control. For instance, the parent task may wait for all
  its sub-tasks to finish before finishing, and conversely all the
  sub-tasks may be killed if the parent task is killed.
\item Pool-agent: specific to the architecture described
  in Section~\ref{sec:pool}. This is an interaction used when agents retrieve
  work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. 
\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable
  for every workflow language. It has been developed mostly for
  well-structured and relatively simple workflow language such as between GWorkflowDL
  and Scufl~\cite{OLAB-09} and between the 4 languages that were involved in the
  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.
\end{enumerate}

@@

See Figure~\ref{archi:service}. The workflow engine is available
externally to the science gateway, in a service. The science gateway
controls the service through a specific interaction (\texttt{d}) that
might be implemented as a web-service call (e.g. RESTful or SOAP), as
a command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
workflow activities, or as a white box that allows for some
interaction with them. The workflow engine is responsible for
controlling the tasks on the infrastructure (\texttt{b}) and for
performing the required data transfers to execute them
(\texttt{c$_2$}). User data is usually managed through the science
gateway (\texttt{c$_1$}), although it might as well be delivered by
the workflow engine directly to the user.

@@

See Figure~\ref{archi:nested}. In nested workflows (Figure~\ref{archi:nested}-Left), an activity of a
\emph{parent} workflow executed by the \emph{parent} engine is itself
a workflow, called \emph{child} workflow, that is executed by a \emph{child} engine. The
parent and child engines might use different workflow description
languages. They might also execute workflows on different
infrastructures. The parent workflow is also called meta-workflow. The
science gateway communicates with the parent engine through
interaction \texttt{*$_1$}. The science gateway also communicate with
the infrastructure to transfer user data through abstract interactions
\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
with the infrastructure through abstract interactions as well,
\texttt{*$_4$} and \texttt{*$_5$}. The parent engine communicates with
the child engine through abstract interaction
\texttt{*$_6$}. Administrator installs workflows in the science
gateway through interaction \texttt{a$_2$}, and installs software on
infrastructures through interactions \texttt{a$_1$}.

@@

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow activities as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Science Gateway for nested workflow execution with
service invocation as represented in Figure~\ref{archi:nested}. The
parent workflow engine is \texttt{WS-PGRADE}, invoked as a service in
the Science Gateway (step \texttt{1} on
Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
Figure~\ref{archi:nested}). Ten different child engines can be used by
nested workflows, invoked through the \texttt{Submission service}
(step \texttt{2}, interaction \texttt{d}). Each of these engines can
submit jobs and transfer data to a distributed computing
infrastructure (\texttt{DCI}, step \texttt{3}, interactions \texttt{b}
and \texttt{c$_2$}). Data interactions (\texttt{c}) and application
porting ones (\texttt{a}) are not represented on
Figure~\ref{fig:shiwa-architecture}.

@@

\emph{Other features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): ability to describe
  meta-workflows from existing workflows.  \todo{More details about
    why this is important?} \texttt{O$_1$}=0 when the feature is
  available, \texttt{O$_1$}=1 otherwise.
\item Fine-grained debugging (\texttt{O$_2$}): availability of
  fine-grained debugging information about workflow activities. When
  such information is available, workflow is seen as a white
  box. Fine-grained information about workflow activity is required to
  properly troubleshoot workflow executions.  \texttt{O$_2$}=0 when
  such information is easy to access, \texttt{O$_2$}=1
  otherwise. \todo{More details about why this is important?}
\end{itemize}
\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing other
features in the architecture.  \todo{Maybe we could revise the
  categories to avoid such a vague ``other features''.}

@@

\paragraph{Other features} Meta-workflows are not available
(\texttt{O$_1$}=1).  Obtaining fine-grained information about workflow
activities is not straightforward since the science gateway has no
knowledge about the workflow topology, and the workflow engine is
integrated as a task (\texttt{O$_2$}=1).

>>>>

In the last decade, the e-Science workflow community has developed
high-level workflow systems to help developers access distributed
infrastructures such as clusters, grids, clouds and web services,
resulting in tools such as Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE/gUSE~\cite{Kacsuk2012}, etc. Such workflow engines usually
describe applications in a specific language that offers parallel data
and control flow constructs, visual edition features, links with
domain-specific tool repositories, provenance recording, etc.
% Descriptions and experiments
%conducted with e-Science workflow engines were published in journals
%such as Future Generation Computer Systems and conference venues such
%as WORKS.

At the same time, specific toolboxes were emerging in various
scientific domains to facilitate the assembly of software components
in consistent pipelines. For instance, in neuroimaging, our primary
domain of interest, tools such as Nipype~\cite{gorgolewski2011nipype},
PSOM~\cite{bellec2012pipeline}, SPM~\cite{ashburner2011spm} and
FSL~\cite{Jenkinson2012782} provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing systems, in particular clusters, to create tasks, handle
their dependencies, and execute them. For instance, FSL can launch
tasks on SGE through its \texttt{fsl\_sub} tool, Nipype has execution
plugins for S/OGE, Torque, IPython, ssh and multi-processor servers,
and PSOM \todo{...}. Some of these tools also support advanced
features such as provenance tracking (Nipype, PSOM), redundancy
detection across analyses to avoid re-computation (PSOM), and so on. A
wide-array of workflows have been implemented using these pipelines
and are now widely shared across neuroimaging groups
world-wide. \todo{Provide more context on that?} These domain-specific
engines represent a tremendous opportunity for science gateways to
leverage existing tools and applications. They nicely complement
e-Science engines that are more oriented towards the exploitation of
distributed computing infrastructures, in particular grids and clouds.

We are looking for software architectures that are able to integrate
e-Science and domain-specific workflow engines in science
gateways. For the sake of the analysis conducted in this paper, we
define a \emph{workflow engine} (also abbreviated \emph{engine}) as a
software that submits interdependent computing \emph{tasks} to an
infrastructure (local server, cluster, grid or cloud) based on a
workflow description (a.k.a \emph{workflow}) and input data that may
be files, data base entries, or simple parameter values. Although
simplistic, this definition covers both e-Science and domain-specific
engines. Some workflow engines, usually e-Science ones, may transfer
data across the infrastructure, and others, usually domain-specific
ones, may leave it to external processes. Workflows may be expressed
in any language, including high-level XML or JSON dialects such as
\texttt{Scufl} or \texttt{Hyperflow}, and low-level scripts such as
\texttt{bash}.  \todo{Mention other workflow definitions,
  e.g. abstract/concrete, etc?}

@@

Several workflow engines may be combined in the same
science gateway.

A workflow
  engine is not supposed to be aware of the science gateway.

@@

@@

\item Workflow integration: consists in adding a new workflow to the
  system so that users can execute it. It is triggered by an
  administrator of the science gateway and it results in an interface,
  for instance a web form, where users can enter the parameters of the
  workflow to be executed. The interaction has two aspects:
  (\texttt{a$_1$}) the programs used in the workflow are installed on
  the infrastructure, which may or may not require administrative
  privileges on the infrastructure; (\texttt{a$_2$}) the workflow is
  configured in the science gateway so that it becomes available to
  users. Note that integrating a workflow is not the same process as
  integrating a workflow \emph{engine}.
\item Task control: operations to manage tasks on the infrastructure,
  including: authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires to deal with the
  heterogeneous batch managers and meta-schedulers that might be
  available on the infrastructure. When the infrastructure is a grid
  or a cloud, it is for instance implemented using libraries such as
  SAGA, DRMAA, OCCI and similar initiatives \todo{refs needed}.
\item Data control: operations to manage data on the infrastructure,
  such as: upload, download, deletion, browsing, replication, caching,
  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer data across the infrastructure. The
  infrastructure might offer various data storage backends with
  heterogeneous interfaces. Tools and services such as
  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
  can be used to homogeneize these interfaces.
\item Workflow control: operations to execute a workflow with an
  engine, including: workflow submission, monitoring, termination,
  etc. Workflow control can be coarse-grained (a.k.a. black box) or
  fine-grained (white box). In a coarse-grained model, the various
  tasks created by a workflow execution are masked and the user only
  has a global view of the workflow execution. In a fine-grained
  model, user is exposed to the workflow topology, i.e. to the outputs
  of the individual tasks, their statuses and so on.
\item Sub-task control: operations used by tasks to submit sub-tasks
  on the infrastructure, including: submission, monitoring,
  termination, deletion, etc. Sub-task control is similar to
  interaction \texttt{b}, except that information about the parent
  task which submitted a sub-task is usually available and used for
  additional control. For instance, the parent task may wait for all
  its sub-tasks to finish before finishing, and conversely all the
  sub-tasks may be killed if the parent task is killed.
\item Pool-agent: specific to the architecture described
  in Section~\ref{sec:pool}. This is an interaction used when agents retrieve
  work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. 
\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable
  for every workflow language. It has been developed mostly for
  well-structured and relatively simple workflow language such as between GWorkflowDL
  and Scufl~\cite{OLAB-09} and between the 4 languages that were involved in the
  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.
\end{enumerate}

@@

See Figure~\ref{archi:service}. The workflow engine is available
externally to the science gateway, in a service. The science gateway
controls the service through a specific interaction (\texttt{d}) that
might be implemented as a web-service call (e.g. RESTful or SOAP), as
a command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
tasks, or as a white box that allows for some interaction with
them. The workflow engine is responsible for controlling the tasks on
the infrastructure (\texttt{b}) and for performing the required data
transfers to execute them (\texttt{c$_2$}). User data is usually
managed through the science gateway (\texttt{c$_1$}), although it
might as well be delivered by the workflow engine directly to the
user.

@@

See Figure~\ref{archi:nested}. In nested workflows
(Figure~\ref{archi:nested}-Left), a task of a \emph{parent} workflow
executed by the \emph{parent} engine is itself a workflow, called
\emph{child} workflow, that is executed by a \emph{child} engine. The
parent and child engines might use different workflow description
languages. They might also execute workflows on different
infrastructures. The parent workflow is also called meta-workflow. The
science gateway communicates with the parent engine through
interaction \texttt{*$_1$}. The science gateway also communicate with
the infrastructure to transfer user data through abstract interactions
\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
with the infrastructure through abstract interactions as well,
\texttt{*$_4$} and \texttt{*$_5$}. The parent engine communicates with
the child engine through abstract interaction
\texttt{*$_6$}. Administrator installs workflows in the science
gateway through interaction \texttt{a$_2$}, and installs software on
infrastructures through interactions \texttt{a$_1$}.

@@

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Science Gateway for nested workflow execution with
service invocation as represented in Figure~\ref{archi:nested}. The
parent workflow engine is \texttt{WS-PGRADE}, invoked as a service in
the Science Gateway (step \texttt{1} on
Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
Figure~\ref{archi:nested}). Ten different child engines can be used by
nested workflows, invoked through the \texttt{Submission service}
(step \texttt{2}, interaction \texttt{d}). Each of these engines can
submit jobs and transfer data to a distributed computing
infrastructure (\texttt{DCI}, step \texttt{3}, interactions \texttt{b}
and \texttt{c$_2$}). Data interactions (\texttt{c}) and application
porting ones (\texttt{a}) are not represented on
Figure~\ref{fig:shiwa-architecture}.

@@

\emph{Other features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): ability to describe
  meta-workflows from existing workflows.  \todo{More details about
    why this is important?} \texttt{O$_1$}=0 when the feature is
  available, \texttt{O$_1$}=1 otherwise.
\item Fine-grained debugging (\texttt{O$_2$}): availability of
  fine-grained debugging information about workflow tasks. When
  such information is available, workflow is seen as a white
  box. Fine-grained information about workflow tasks is required to
  properly troubleshoot workflow executions.  \texttt{O$_2$}=0 when
  such information is easy to access, \texttt{O$_2$}=1
  otherwise. \todo{More details about why this is important?}
\end{itemize}
\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing other
features in the architecture.  \todo{Maybe we could revise the
  categories to avoid such a vague ``other features''.}

@@

\paragraph{Other features} Meta-workflows are not available
(\texttt{O$_1$}=1).  Obtaining fine-grained information about workflow
tasks is not straightforward since the science gateway has no
knowledge about the workflow topology, and the workflow engine is
integrated as a task (\texttt{O$_2$}=1).
